Fail Faster

### 2025-03-11 paper/paper.tex
some updates

\definecolor{dkred}{rgb}{0.7,0,0}
\definecolor{dkblue}{rgb}{0,0,0.7}
\definecolor{notered}{rgb}{0.85,0,0}
\definecolor{dkpurple}{HTML}{4e02eb}
\definecolor{dkgreen}{HTML}{006329}
\definecolor{teal}{HTML}{007982}
\definecolor{fuchsia}{HTML}{8C368C}

\newcommand{\comm}[3]{\ifdraft\textcolor{#1}{[#2: #3]}\fi}
\newcommand{\cn}{\ifdraft{\color{blue} [CITE]}\fi}
\newcommand{\hg}[1]{\comm{dkred}{HG}{#1}}
\newcommand{\tr}[1]{\comm{dkblue}{TR}{#1}}
\newcommand{\bcp}[1]{\comm{dkpurple}{BCP}{#1}}
\newcommand{\jwc}[1]{\comm{dkgreen}{JWC}{#1}}
\newcommand{\outline}[1]{\ifdraft{\color{teal} #1}\fi}

\newcommand{\bq}{\texttt{base\_quickcheck}}
\newcommand{\name}{\texttt{waffle\_house}}
\newcommand{\tool}{Waffle House}

\newcommand{\secref}[1]{\S\ref{#1}}

\newcommand{\proposecut}[1]{\ifdraft{\color{gray} [#1]}\fi}
\newcommand{\proposechange}[2]{\ifdraft{\color{gray} {#1}
    $\rightarrow$ #2}\else #2\fi}

@@

\author{Joseph W. Cutler}
\affiliation{%
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}}

@@

\author{Benjamin C. Pierce}
\affiliation{%
 \institution{University of Pennsylvania}
 \city{Philadelphia}
 \state{Pennsylvania}
 \country{USA}}
%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Richey et al.}

@@

\section{Introduction}
Property-based testing (PBT) is a widely-used testing framework consisting of
two key components: a set of \textit{properties} that a system must satisfy, and
a large number of \textit{inputs} to that system. In contrast to traditional
unit testing, where inputs are hand-written, PBT users create a random-input
\textit{generator} that automatically produces inputs. Sometimes, this process
can itself be automated through the use of PBT libraries that synthesize
generators from type definitions; however, when inputs are required to maintain
invariants not communicated by the type, users must write a custom generator (or
filter out a broad class of invalid inputs).

Writing ``good" generators—those that uncover more bugs in the system under
test—is challenging. Much effort has been dedicated to this problem, often by
developing sophisticated domain-specific languages for specifying constrained
generators. However, these approaches do not always reflect the needs of real
PBT users. Under ideal conditions, a PBT suite would continue to run
until it stopped finding bugs, real-world usage patterns are very different; a
study of expert PBT users reported time budgets of between 50 milliseconds and
30 seconds for their test suites~\cite{inpractice}. Consequently, generation
time is a significant factor in PBT efficacy---that is, if a generator can
produce inputs twice as fast, it has twice as many chances within a given time
bound to find a bug. Therefore, it is imperative that PBT libraries are built
with efficiency in mind.

@@

In summary, we dramatically improve the bug-finding effectiveness of PBT generators
by optimizing them for speed. In speicific, we make the following contributions:
\begin{enumerate}
    \item An empirical analysis of the sources of inefficiency in PBT
    generators. 
    \item A library, \name, which offers efficient generator functions through the
    use of staged metaprogramming; this is the first known application of metaprogramming to PBT.
    \item The insight that generators should be compared by program equality 
    \item An evaluation demonstrating the improved performance of generators constructed
    using our library in a controlled comparison.

@@

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, splice, stage distinction.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
    \item We can convert one way, but not the other.
    \item Functions \texttt{'a code -> 'b code} ``fuse''.
    \begin{itemize}
      \item Consider writing \texttt{even . succ}. This
      \item Consider \texttt{even_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ_c : int code -> int code = fun cx -> .< .~x + 1 >.}
      \item If you do \texttt{to_dyn (even_c . succ_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
      a dynamic function value ensures that the resulting code is free of abstractions.
    \end{itemize}
    and 
  \end{itemize}
}

@@

@@

\tr{Discuss BQ, including type-derived generators.}

@@

\jwc{We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}

\jwc{
  \begin{itemize}
    \item How does (simplified) BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show benchmarks of the running example, versus the version where you inline everything. ()
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

@@

@@

\jwc{
  \begin{itemize}
    \item In cases where the compiler cannot statically eliminate them, binds allocate short-lived closures: we allocate a closure for the continuation, and then immediately jump into it.
    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast, and each minor allocation brings us closer to the next costly GC pause.
  \end{itemize}
}

@@

\jwc{
  \begin{itemize}
    \item The pRNG at the heart of BQ is the hottest part of the hot path: generating a single int with \texttt{int st lo hi} samples from the underlying rng many many times.
    \item BQ's prng is the SplitMix algorithm, implemented in the splittablerandom package.
    \item The precise details of how SplitMix works are unimportant for this paper, but the core of it does an enormous amount of 64-bit int arithmetic.
    \item Because of implementation details related to the garbage collector, OCaml's \texttt{int} type is 63 bits, so an \texttt{Int64.t} is actually \emph{boxed}: a pointer to a custom block, not scanned by the garbage collector.
    \item Because of this, all \texttt{Int64.t} operations (both arithmetic and bitwise operations) must allocate! A single call to \texttt{int} in Bq calls the samples the RNG many times, and each sample (call to \texttt{next\_int64}) allocates 9 words (2 multiplications, one addition, and 6 logical operations).
  \end{itemize}
}

\jwc{We discuss ways to fix this in section ...}

\jwc{We note that this inefficiency is not due to the ``abstraction overhead'' of the DSL itself, but simply due to language implementation and algorithm choices.
We include this to demonstrate the amount of slack in the system, and so we can demonstrate the additional testing-time savings in the evaluation section.
}

@@

\jwc{
  \begin{itemize}
    \item The simplest combinators are presented below.
    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
    \texttt{code} for the result of running the first generator. At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
    \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
    The function application \texttt{g size_c random_c} runs during the first stage, producing the code for the \texttt{a}. This is then spliced into the quote: the specialized code of the generator \texttt{g}
    put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
    \item \texttt{}
  \end{itemize}
}

@@

\subsection{Faster Random Number Generators}

@@

>>>>

\include{macros}

@@

\author{Joseph W. Cutler}
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \state{Pennsylvania}
  \country{USA}
}

@@

\author{Benjamin C. Pierce}
\affiliation{%
 \institution{University of Pennsylvania}
 \city{Philadelphia}
 \state{Pennsylvania}
 \country{USA}}
%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Richey and Cutler et al.}

@@

\section{Introduction}

\jwc{This paper is about ``Strict functional languages like OCaml and Scala'', but we'll focus on OCaml for presentation.}

Property-based testing (PBT) is a widely-used testing framework consisting of
two key components: a set of \textit{properties} that a system must satisfy, and
a large number of \textit{inputs} to that system. In contrast to traditional
unit testing, where inputs are hand-written, PBT users create a random-input
\textit{generator} that automatically produces inputs. Sometimes, this process
can itself be automated through the use of PBT libraries that synthesize
generators from type definitions; however, when inputs are required to maintain
invariants not communicated by the type, users must write a custom generator (or
filter out a broad class of invalid inputs).

Writing ``good" generators—those that uncover more bugs in the system under
test—is challenging. Much effort has been dedicated to this problem, often by
developing sophisticated domain-specific languages for specifying constrained
generators. However, these approaches do not always reflect the needs of real
PBT users. Under ideal conditions, a PBT suite would continue to run
until it stopped finding bugs, real-world usage patterns are very different; a
study of expert PBT users reported time budgets of between 50 milliseconds and
30 seconds for their test suites~\cite{inpractice}. Consequently, generation
time is a significant factor in PBT efficacy---that is, if a generator can
produce inputs twice as fast, it has twice as many chances within a given time
bound to find a bug. Therefore, it is imperative that PBT libraries are built
with performance in mind.

@@

In summary, we dramatically improve the bug-finding effectiveness of PBT generators
by optimizing them for speed. In speicific, we make the following contributions:
\begin{enumerate}
    \item An empirical analysis of the sources of inefficiency in PBT
    generators. \jwc{And an argument that PBT generator libraries should be engineered for performance}
    \item A library, \name, which offers efficient \jwc{performant?} generator \proposechange{functions}{combinators} through the
    use of staged metaprogramming; this is the first known application of metaprogramming to PBT.
    \item The insight that generators should be compared by program equality 
    \item An evaluation demonstrating the improved performance of generators constructed
    using our library in a controlled comparison.

@@

\jwc{For the purposes of this paper, we will only use two stages: compile time, and run time.}

\jwc{
  \begin{itemize}
    \item The \texttt{'a code} type, quote, splice, stage distinction.
    \item The difference between \texttt{('a -> 'b) code} and \texttt{'a code -> 'b code}
    \item We can convert one way, but not the other.
    \item Functions \texttt{'a code -> 'b code} ``fuse''.
    \begin{itemize}
      \item Consider writing \texttt{even . succ}. This
      \item Consider \texttt{even\_c : int code -> bool code = fun cx -> .<.~cx mod 2 == 0>.} and \texttt{succ\_c : int code -> int code = fun cx -> .< .~x + 1 >.}
      \item If you do \texttt{to\_dyn (even\_c . succ\_c)}, you get \texttt{fun x -> (x + 1) mod 2 == 0}. Composing functions from code to code, and then \emph{only at the end} stamping out
      a dynamic function value eliminates the function abstraction.
    \end{itemize}
    \item This is the basis of (WORD). By defining a library with functions with types like \texttt{'a code -> 'b code}, we can ensure that the
    abstractions the library introduces are fully eliminated at compile time.
  \end{itemize}
}

@@

\jwc{TODO: Move this section later (cf conversation in WH meeting 3/7/25)}

@@

@@

\jwc{NOTE: We can simplify this further by getting rid of the size parameter. Make the story even cleaner, I think!}

\jwc{
  \begin{itemize}
    \item It's been long-known that clean functional abstractions have a runtime overhead (this should be familiar by this time in the paper).
    \item How does (simplified) BQ work?
    \begin{itemize}
      \item The basic generator type: \texttt{'a generator = int -> SR.t -> 'a}. Size and random seed to deterministic value. (note: \texttt{SR.t} is a mutable seed)
      \item This gets a monad intance in the obvious way (show code).
      \item Also show the code for \texttt{int}, how it calls the underlying SR function.
    \end{itemize}
    % \item Note that \emph{extensionally} \texttt{generate (create (fun size random -> e)) size random = e}, but the OCaml compiler does not always perform this optimization, or do the inlining required to expose it.
    % (When sufficiently obfuscated behind returns and binds ...) This program compiles to code that (1) allocates the closure for `e', (2) passes it to create (which returns the closure), and then calls (3) generate, which immediately jumps into the closure.
    \item Show benchmarks of the running example, versus the version where you inline everything. ()
    \item Let's look at the running example: inline it all the way.
  \end{itemize}
}

@@

\jwc{
The problem is plain: while the abstractions that PBT libraries like base quickcheck provide are indespensible for writing idiomatic generators, the performance overhead of using them is dramatic.
The compiler (and flambda) uses heuristics (\href{https://ocaml.org/manual/5.3/flambda.html}{here}) to decide when to specialize and inline code, and these heuristics cannot guarantee that generator abstractions are zero-cost.
These heuristics are also extremely conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
}

@@

\jwc{
  \begin{itemize}
    \item While it's the ``correct'' abstraction for generators, using a monadic interface prevents the compiler from specializing generator code. Using monadic bind and return obscures the control and data flow of a generator from the compiler. This is
    compounded when handling recursive functions.
    \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.
    \item Each closure allocation is relatively cheap (they're going to be minor allocations since they're short-lived), but doing lots of allocation in a generation hot loop adds up fast, and each minor allocation brings us closer to the next costly GC pause.
  \end{itemize}
}

@@

The core of any PBT generator library is a random-number generator. Following
the original Haskell QuickCheck, \bq uses the SplitMix algorithm \cn,
implemented as an OCaml library called \texttt{Splittable\_random}. The precise
details of how SplitMix works are unimportant for the present paper, but the
main operations are defined in terms of bitwise operations on 64-bit integers.
However, due to implementation details related to the garbage collector, all
runtime OCaml values are either (a) pointers to a block of memory, or (b) 63-bit ``immediate'' values.
As such, the OCaml type of 64-bit ingeters, \texttt{int64}, is \emph{boxed}: it
is a pointer to a block of memory that contains a 64-bit integer.  This means
that \emph{all} \texttt{int64} operations (both arithmetic and bitwise) must
allocate memory cells to contain their output.

This has a significant impact on the performance of generators. A single call to
one of the \bq library functions may sample from \texttt{Splittable\_random}
multiple times, and each sample from \texttt{Splittable\_random} allocates 9
words \jwc{this is a call to \texttt{next\_int64}}.  While seemingly small, we
will see in Section~\ref{section:eval} that this amount of allocation has a
significant impact on significant.

In Section~\ref{subsection:faster-rng}, we show how our library mitigates this
performance hit by using OCaml's FFI to reimplement the RNG in non-allocating C code.
We also discuss how the ``unboxed types'' extension to the OCaml compiler \cn could instead be used
to the same effect without leaving OCaml.

It is worth noting that the inefficiency of boxed \texttt{int64}s is not due to
the abstraction overhead of the generator DSL itself, but simply due to language
implementation and algorithm choices. We include this discussion --- and
the corresponding evaluation
--- to demonstrate the testing-time savings of taking seriously the notion that PBT libraries
should be engineered for performance.

@@

\jwc{
  \begin{itemize}
    \item The simplest combinators are presented below.
    \item \texttt{return} is the constant generator, given code for \texttt{x : 'a}, it gives the generator which always generates \texttt{x}
    \item \texttt{bind} is like BQ's bind --- sequencing generators by passing the result of one to another --- except that the continuation gets access to the
    \texttt{code} for the result of running the first generator. At compile time, we know only that \texttt{g} will return \emph{some} \texttt{a}, but we do not get access to the particular \texttt{a}.
    \item Bind takes code for the size and seed, and returns code that first let-binds the result of the first generator.
    The function application \texttt{g size\_c random\_c} runs during the first stage, producing the code for the \texttt{a}. This is then spliced into the quote: the specialized code of the generator \texttt{g}
    put on the right-hand side of the let. Then, we splice in the code for the continuation, passing first the variable x, and the same size and random seed.
  \end{itemize}
}

@@

\subsection{Faster Random Number Generators}
\label{subsection:faster-rng}

@@

\jwc{All of this can be done with Jane Street bleeding-edge compiler extensions, though these are not yet compatible with MetaOCaml.}
